A Bayesian approach to burial networks to infer social changes in northeastern Taiwan during the European colonization period

### 2021-02-08 analysis/paper/paper.Rmd
reorganize the Intro and the background sections

<!--burial study- network analysis-->
Burial analysis by archaeologists is an approach to understand past social structures through the study of the physical traces of mortuary practices. The material culture and biological records of burial behaviors can represent the social ranking or identities of the deceased [@Binford1971; @Drennan2010; @Saxe1970]. Despite criticism that manipulation of burial rituals by the living can cause a disconnect between a person's status in life and their status represented by burial contexts [@Hodder1980; @Pearson1982], burials can still provide valuable information to infer past societies [@Chapman2003]. By studying burial forms, grave goods, and ritual behaviors that structure the material configuration of burials, we can infer social status and relations between members of a community that give an insight to social structures [@Byrd1995; @Seikel2011]. Based on network science and graph theory, the recent development of social network analysis provides many new tools to visualize and analyze relationships in archaeological data [@Borgatti2009; @Brughmans2013]. Network science is the study of relational data where a phenomenon is conceptualized as a network through two steps, abstraction and representation [@Brandes2013; @Collar2015]. A network can be viewed as a patterned aggregation that includes individual elements (i.e. individual burials), pair-wise relationships (the dyads, for example burials with similar types and amounts of grave goods), and an overall structure showing global patterns represented in the data. Burials with the same kinds of prestige goods could indicate some underlying social relations where individuals share similar access to trade, exchange, and gifting networks according to their status [eg. @Coward2013, pp. 252]. This paper will use a novel Bayesian approach on exponential random graph models for networks to investigate the formation of relationships between burials in northeastern Taiwan to explore the indirect impact of foreign presence on social organization.

<!-- intro to networks: move to SI -->

<!-- two types of network analysis: move to SI -->

@@

<!-- archaeological background -->
Archaeological sites in northeastern Taiwan show the evidence of accumulation patterns of imported goods, such as ornaments and porcelains, in residential areas and burial contexts around the same time as European contacts in the 17th century [@Chen2007; @LWandBM2020ornament]. Imported goods were treated as prestige or wealth in those Indigenous societies that were commonly mentioned in Spanish historical records in the 17th century (1624-1642) [@Borao2009; @LiandWu2006]. For example, a document written by a Spanish priest described a Spanish soldier using carnelian beads to exchange for natural resources with local Indigenous people, because of their high value in Indigenous culture [@LiandWu2006]. In addition, Spanish visitors observed that an Indigenous person possessing more imported goods may have been recognized as more influential or having higher status in their community [@LiandWu2006]. Despite the fact that we recorded an uneven distribution of imported goods in burials, the degree of differentiation across burials over time remained unclear. The introduction of foreign prestige goods to local Indigenous communities might have occurred before European presence became established, but was maintained and amplified after European arrival due to local competition for power [@Borao2009; @LiandWu2006; @LWandBM2020ornament; @Ueda2016]. A general model to summarize this situation in northeastern Taiwan might assume that the influence of a colonial power, combined with high local values attached to imported goods, might lead to increased social inequality due to competition among individuals [@Brumfiel1994; @Clark1994]. The observed uneven distribution in Northeastern Taiwan may be explained as a result of unequal access to trade goods when Indigenous societies became more involved in the complex trade network stimulated by Europeans.

<!-- archaeological specific case study -->
In what follows, we explore social changes in a pericolonial context at Kiwulan (Figure \@ref(fig:figure-KWL-map)), an Iron Age site in northeastern Taiwan, which was occupied from 14th to 19th centuries. This occupation period includes the time before the European arrival, the presence of the Europeans, including the Spanish and the Dutch, in the 17th century, and finally a large wave of Han Chinese in the 19th century [@Chen2007]. We argue that the involvement in the long-distance trade stimulated by the Europeans might lead to increased social inequality at Kiwulan. This argument would be supported by changes in the structure of networks built on burial data based on the assumption that burials can reflect social structure. We examine burial networks before and after the foreign contacts, and test our hypothesis that increased social inequality can be observed in the network after the European arrival in 17th century Taiwan. We ask: (1) did European colonial activities result in increased social inequality in Indigenous society in ways that can be detected by analysis of burial networks? (2) what are the major variables affecting or forming the unequal social positions represented by higher degree of network popularity (e.g., a few nodes have more relationships with other nodes) that might hint at social heterogeneity influence by the European arrival? By answering these questions, this case study helps to expand our understanding of European colonial effects on Indigenous groups in an indirect way by examining burial data. We also introduces a novel Bayesian approach on statistical modeling for networks that is suitable for networks with smaller sizes.

@@

# Exponential random graph models in a Baysesian framework  {-}

<!-- ERGM in archaeology, and limitations --> 
Exponential random graph models (ERGMs) first appeared in archaeology with @Brughmans2014, who studied Iron age settlement patterns in Southern Spain by modeling inter-settlement visibility networks and visual control at 159 sites. Although their models do not fit perfectly with the observed network of archaeological data, they proposed that ERGM is a promising method for exploration and hypothesis testing for social processes. Similarly, @Amati2019 modeled three networks consisting of 15 sites (AD 100 to 400) in the Caribbean to explore interaction mechanisms, including proximity, inter-cultural items, and pottery types. By comparing hypothesized networks with the observed sites, they found that the presence of hub sites can be efficiently explained by multiple interdependent mechanisms instead of only one variable exclusively. However, those studies also point out some limitations of ERGM, such as sensitivity to missing data and less ability to handle uncertainty. In addition, it is difficult in ERGM to estimate model parameters and interpret the result due to intractable likelihood normalizing constants and model degeneracy [@Caimo2014; @Jin2013]. A normalizing constant is a function of the model parameter for making probability distributions integrate to one, which becomes harder to compute with a larger set of networks [@Caimo2017bayesian]. This is also termed "doubly intractable" since both the likelihood normalizing constant and the marginal likelihood (the evidence of the posterior) are hard to derive [@Caimo2013; @Lyne2015]. Model degeneracy is another issue where probability models tend to overestimate a small number of extreme graphs by assigning too much weight, e.g., in extreme cases such as empty (all nodes unconnected) or complete graphs (all nodes connected) [@Schweinberger2011; @Caimo2014]. One solution to these limitations is available by implementing ERGMs by means of a Bayesian framework.

@@

<!-- Bayesian inference for Exponential Random Graph Models --> 
Bayesian approaches to ERGMs are effective tools for network modeling by incorporating prior information about the network configurations to better understand dependencies of network variables and improve computational issues in ERGMs [@Caimo2017; @Lehmann2020]. Prior information is derived from previous data or assumptions of our data. The advantage that Bayesian modeling has over traditional ERGMs is the application of Markov chain Monte Carlo (MCMC) simulation using the approximate exchange algorithm [@Caimo2011]. With the exchange algorithm, Bayesian ERGMs avoid doubly-intractable computations by directly sampling from the not normalized part of the posterior, which alleviates the computational problems and gives better convergence results. This enables us to deal with complicated dependence patterns with ease, providing better estimations for complex social network models with heterogeneous data [@Caimo2017; @Snijders2006]. By fitting an ERMG with an MCMC algorithm, a Bayesian approach generates posterior probabilities that incorporate our sample data and prior information that estimates the effect of each ERGM parameter in our models [@Caimo2015; @Nemmers2019]. This allows uncertainty quantification by examining the posterior mean and 95% credible intervals, or the odds ratios (conditional probability of observing an edge) that can replace p-values for assessing a null hypothesis [@Caimo2017bayesian]. The typical criteria for interpreting an significant effect of a parameter is that its credible interval does not contain zero, or an odds ratio is greater/less than one, meaning a significant positive/negative effect [@Harris2013, pp. 51]. In addition, Bayesian approaches are also useful to deal with missing data, which is often a problem leading to misinterpretation of networks, especially for archaeological studies. @Koskinen2010 shows that the effect of missing data can be reduced with Bayesian modeling that can predict, on the average, 80% of the ties when a third of data is missing.

@@

<!-- archaeological setting -->
We analyzed burial data collected from the published excavation reports, and the original fieldwork notes for the upper component of Kiwulan, an Iron Age settlement (1350-1850 AD) [@Chen2007]. Excavations revealed abundant trade goods, including imported ceramics and stonewares, metal artifacts, glass beads, and stone beads, indicative of frequent trade activities [@Chen2007]. A total of 90 burials were unearthed that are mostly located in the middle section of the excavation area, which is the largest open area at Kiwulan that provides continuous stratigraphic sections suitable for temporal comparison (Figure \@ref(fig:figure-burial-map)). Burials are oriented in an east-west direction on the north side of the residential area, indicated by post-holes and _in-situ_ wooden posts, suggesting a well-organized spatial arrangement. Previous studies report an uneven distribution of prestige goods across burials without agreement about whether this uneven distribution hints at vertical social differences. For example, @Cheng2008 interpreted the unequal distribution of glass beads, especially the gold-foil beads between burials, as evidence for hierarchy, indicating a stratified society. However, @Hsieh2012 suggested a relatively egalitarian structure based on comparative analysis of the frequencies of all burial goods. She found that the burials with rare goods, also prestige goods in local culture, were usually associated with elders, which might indicate achieved, rather than inherited, status. One important limitation of these previous studies is that they did not use analytical units suitable for comparing the phases before and after European arrival. Here, we adopt a new chronological framework for the burials to test if network configurations differ from the pre-European period (before European colonization) to the post-European period (during and after European colonization until Han Chinese immigrants arrival). The discord over the discussion of Kiwulan social organization could be associated with chronological differences that have not yet been well studied.

@@

<!-- burial data and the definition of tie formation  --> 
We used burial goods in common to construct ties for representing a flow of goods between burials. Since trade beads are commonly found across burials with substantial differences in quantities, we described each burial as having one of four levels, high, upper-middle, lower-middle, and low, according to their distributions across all burials (details in Supplementary Online Materials). For less frequent prestige goods, including imported ceramics, gold foils, and fish-shaped ornaments, we linked two burials when they possess each type of goods (i.e. presence or absence). Node attributes here include osteological data, such as age and sex, and cultural data, such as ritual pottery, and burial value index. Ritual pottery was identified as locally made ceramics placed above graves that suggests funeral feasting based on historical records. The burial value index is an important attribute for economical inequality that we assigned the burials into three ranks, high (>`r high_value`, top 10 percent), above average (`r ave_value`-`r high_value`), and below average (<`r ave_value`), as an index of wealth. The number is the sum of values of each type of prestige good, which is calculated by the total number of prestige item from burial context over the number of prestige items in a burial. Since burials tend to have multiple prestige goods in common, the network ties are weighted instead of binary data (the value 1 represents a tie and the value 0 otherwise) [@Snijders2011]. For example, if two burials have both low quantity of glass beads and porcelain in common, the tie is given a value of 2. Our networks are non-directed, which means a mutual relationship where the tie between any two actors is bidirectional. The networks constructed based on these principles show that the network after the European presence has more node connectivity in general with some nodes having a larger number of connections (Figure \@ref(fig:figure-network-diagrams)).

@@

<!-- Network construction and specification -->
We quantify the relations among burials and test our hypothesis of social change using the R programming language [@Rlanguage2019] with the bergm package [@Caimo2014], where model specification and simulation procedures is based on statnet package [@Hunter2008; @Morris2008; @Handcock2008]. Table \@ref(tab:table-ergm-terms) lists the network parameters we used for the model specification with the corresponding archaeological evidence. Every parameter in an ERGM has an associated algorithm for computing the probability of observing relations in terms of grave goods between two burials. Based on our hypothesis, we model a network with increased social inequality to be represented by endogenous network effects: low transitivity and high popularity. We include burial-specific attributes as covariate effects for homophily, such as age, sex, ritual activity, and the degree of wealth, to explore the tendency of tie forming. For example, age-homophily means people of the same age tend to have the same burial goods. We also include the physical distance between burials as an indicator of a kinship-based relation since the deceased from the same family tend to be buried nearby [@LiandWu2006]. Our model could reveal the emergence of social inequality via the presence of a few individuals as network centers, having more relations with others. This would be indicated by high popularity or degree values in the network statistics, with covariates to understand the preferential tendency of formation of relationships. According to our hypothesis, the burial evidence from after the European arrival will show higher popularity compared with the burial evidence from before European contact.

@@

<!--posterior stats for for post-E and their comparison-->
For the network model after the arrival of Europeans, the nodal covariates of wealth rank show significant positive effects, while ritual and sex variables have negative effects. This indicates the burials in the same wealth level tend to have relations. On the contrary, the same sex burials and burial with ritual pottery in common tend not to have relations. There are no significant effects for variables of age and physical distance. Similar to the pre-European network, the endogenous network variable, transitivity (gwesp) demonstrates a significant positive effect but much lower than the effect of pre-European network. In contrast, the popularity (gwdegree) shows a significantly higher positive effect than the effect of the pre-European network. This means there is a tendency toward centralization that reflects a limited number of burials having many more ties than others. This implies that the presence of a better access to trade goods and the behavior of wealth accumulation or display in burial events. In general, the post-European network model has a smaller transitivity effect and a positive popularity effect than the pre-European network model. This may suggest a reduced tendency toward clustering but high tendency toward centralization after the European presence. In addtion, both posterior estimates present symmetric distributions, with the posterior means close to posterior medians (Table \@ref(tab:table-posterior-stats)).

@@

<!-- comparison between two networks using bootstrap --> 
One key difference between the pre-European and post-European networks is their size, with `r pre_E` burials compared to `r post_E` burials. To understand the robustness of comparison between two networks, we used vertex bootstrap to cross-validate the results of Bayesian ERGMs. Vertex bootstrap is a non-parametric method that conducts resampling for all vertices (i.e. node) that is useful for quantifying standard errors and estimating sampling variability in the network statistics of interest [@Chen2019; @Snijders1999; @Rroberts2021]. This enables the evaluation of uncertainty for networks and tests the difference between multiple networks by examining their confidence intervals for the network population. Networks with nonoverlapped intervals means there is a significant difference. We computed and compared endogenous network statistics, including density, popularity, and transitivity for our two networks. Figure \@ref(fig:figure-bootstrap-CI) shows a significant difference in observed network popularity between the two networks, which is consistent with our finding of negative popularity in the pre-European period and positive popularity in the post-European period using the Bayesian approach. For density and transitivity, the 95% confidence intervals from two networks overlapped with each other that indicates no significant difference. This is also consistent with our results of Bayesian modeling, where both networks present similar positive or negative effects with some degree of difference.

@@

<!--archaeological implication based on the results-->
The covariants for networks provide further information to give insights into how specific elements of the archaeological record relate to the formation of networks (Table \@ref(tab:table-posterior-stats)). In other words, we can know what kinds of socio-cultural factors associated with the acquisition and distribution of foreign prestige goods reflected by burial networks. Our covariants were age, sex, ritual pottery, and wealth ranks calculated based on the value of grave goods. The results indicate that the major factor contributing to ties between burials is ritual for the pre-European network, and wealth for the post-European network. Moreover, in the post-European network, burials with ritual vessels or same sex tend not to be tied that suggest they are not associated with the network pattern we observed. Those results illustrate that different social and economic mechanisms determined relations between burials, or the interconnections of prestige goods, for the two periods. Before European arrival, ritual behaviors seem to have been important factors that structure the interconnections of prestige goods in the burial context. This might imply that ritual could be a status indicator when the foreign goods were first introduced to the local society. After European presence was established, status indicators shifted from ritual to wealth differences with the least sex status discrimination. Moreover, geographical proximity, implying some kinship association, is not a factor for sharing similar prestige items between burials for both phases. This may suggest that the use of prestige items is more individual-oriented instead of kinship-based. Demographic information, such as age and sex, do not have any effects on network formation for the pre-European network and only sex shows a negative effect for the post-European network. This may also support a shift to more wealth and individual based status competition without sex differences in the later phase. But we also note that no effects may be due to the unavailability of age and sex values for many burials in our sample.

<!-- other evidence -->
We used burial data as a proxy to explore social relations reflected by interconnections of prestige goods based on the assumption that burials can represent social structures [@Binford1971; @Saxe1970]. We argue that the social relation at Kiwulan changed because of the contact with European colonial power and associated foreign trade. This argument is supported by the changes in burial network structures from before and after European arrival and their associated underlying mechanisms. It should be noted that there could be an issue of disconnection between the role of the deceased in a burial context and their life. However, this issue can reduced by comparing the results with evidence from other archaeological contexts, such as residential areas [@Chapman2003]. Previous studies of trade ornaments from the residential area of Kiwulan suggests an uneven spatial distribution when the Europeans were active in northern Taiwan that hints at an increased social differentiation at Kiwulan in a pericolonial situation [@LWandBM2020ornament]. This is consistent with the result of a more centralized burial network after the presence of Europeans, which supports a connection between a burial and the social role of the living person. Since the burial goods used for making ties were treated as prestige goods or status items in the local Indigenous culture [@Borao2009; @LiandWu2006], the interrelations represented by the flow of goods observed in the burial data likely reflect social contexts of the living people at Kiwulan in a pericolonial context.

@@

We would like to thank the Yilan County Cultural Affairs Bureau in Taiwan for permitting access to the data used in this paper. We thank the excavation team of Kiwulan led by Yu-pei Chen for their post-excavation analysis and detailed documentation about burials that enable this study. We also thank Alberto Caimo and Steven M. Goodreau for the discussion regarding the methods at the earliest stage of analysis. This work was supported in part by the travel grant from the Department of Anthropology at University of Washington and the Doctoral Dissertation Fellowship from the Chiang Ching-kuo Foundation for International Scholarly Exchange (Project #DF009-A-18). We thank Ben Fitzhugh and Peter Lape for their insightful comments on an early draft. We also acknowledge comments from anonymous reviewers that have greatly improved this manuscript.

>>>>

<!--study social changes in pericolonial context using burial data as a proxy-->
Social changes in Indigenous societies when faced with colonial powers are commonly observed in many parts of the world, especially for European colonies where Indigenous economic, culture, or socio-political aspects were substantially impacted [@Dietler2005; @Silliman2005; @Voss2005]. Recent studies demonstrate that the indirect effect of colonialism or involvement in long-distance trade could also have impacts on Indigenous societies in a pericolonial context [@Acabado2017; @Trabert2017; @LWandBM2020ornament]. To explore social changes, burials are important evidence to understand past societies because material culture and biological records of burial behaviors can represent the social ranking or identities of the deceased, and social relations between them [@Binford1971; @Drennan2010; @Saxe1970]. Burial treatments can further imply social complexity or social inequality where foreign goods from long-distance trade were often used by Indigenous people to express status [@Carter2015; @Dolfini2019]. Usually, burial studies are based on characterizing each physical trace or burial variable, such as biological records, grave forms and goods, or ritual behaviors, and finally combining those individual observations to infer past societies [@Byrd1995; @Seikel2011]. Recent development of network modeling provides new tools to explore the structure of the material configuration in archaeological data and dependencies between variables to understand past relationships, interactions, and underlying mechanisms [@Brughmans2013; @Brughmans2018; @Crabtree2015; @Erik2015; @Freeman2004; @Salvini2010; @Ostborn2015].

<!-- intro to networks: move to SI -->
<!-- two types of network analysis: move to SI -->

<!---network analysis and the basic concepts-->
Different from a static characterization of burial variables, network modeling approaches have the potential to study social changes in a dynamic way by examining the intercorrelations between burials. Drawing largely on network science and graph theory, a key assumption in the interpretation of archaeological networks is that a phenomenon can be conceptualized as a network through abstraction and representation [@Brandes2013; @Collar2015; @Mills2017; @Peeples2019]. A network is viewed as a patterned aggregation that includes individual elements (i.e. individual burials), pair-wise relationships (the dyads, for example burials with similar types and amounts of grave goods), and an overall structure showing global patterns represented in the data. Burials with the same kinds of prestige goods could indicate some underlying social relations where individuals share similar access to trade, exchange, and gifting networks according to their status [eg. @Coward2013, pp. 252]. Among the network modeling methods, exponential random graph models (ERGMs) are one important family of statistical graphs to explore complex networks [@Ghafouri2020]. However, the computational difficulties and the sensitivity to uncertainties to some extent limits its applications [@Caimo2014]. In this paper, we try to go beyond ERGMs with a novel Bayesian approach to investigate the formation of material connections between burials and changes in network structures by examining evidence from a case study in northeastern Taiwan.

<!--research aim and questions with archaeological specific case study-->
We study social changes in a pericolonial context at Kiwulan, an Iron Age site in northeastern Taiwan, which was occupied from the 14th to 19th centuries. This occupation period includes the time before the European arrival, the presence of the Spanish and the Dutch in the 17th century, and finally a large wave of Han Chinese in the 19th century [@Chen2007]. We argue that the involvement in the long-distance trade stimulated by the Europeans might lead to increased social inequality at Kiwulan based on the theoretical framework of corporate-network strategies [@Blanton1996; @Feinman2000]. We test our hypothesis that changes in power structure from a corporate mode to a network mode can be observed in the burial network after the European arrival in 17th century Taiwan. We ask: (1) did European colonial activities result in increased social inequality in Indigenous society in ways that can be detected by analysis of burial networks? (2) if so, what are the major variables affecting or forming the unequal social positions represented that might hint at social heterogeneity influence by the European arrival? By answering these questions, this case study helps to expand our understanding of European colonial effects on Indigenous groups in a pericolonial context. In addition, we introduce a novel Bayesian inference on statistical modeling for networks that is suitable for networks with smaller sample sizes.

# Background {-}

@@

<!-- archaeological background in NE Taiwan in general -->
Archaeological sites in northeastern Taiwan show the evidence of accumulation patterns of imported goods, such as ornaments and porcelains, in residential areas and burial contexts around the same time as European contacts in the 17th century [@Chen2007; @LWandBM2020ornament]. Imported goods were treated as prestige or wealth in those Indigenous societies that were commonly mentioned in Spanish historical records in the 17th century (1624-1642) [@Borao2009; @LiandWu2006]. For example, a document written by a Spanish priest described a Spanish soldier using carnelian beads to exchange for natural resources with local Indigenous people, because of their high value in Indigenous culture [@LiandWu2006]. In addition, Spanish visitors observed that an Indigenous person possessing more imported goods may have been recognized as more influential or having higher status in their community [@LiandWu2006]. Despite the fact that we recorded an uneven distribution of imported goods in burials, the degree of differentiation across burials over time remained unclear.

<!--specific case from Kiwulan-->
To explore indirect colonial impacts on Indigenous societies in northwestern Taiwan, our case study focuses on Kiwulan (Figure \@ref(fig:figure-KWL-map)), an Iron Age settlement (1350-1850 AD) [@Chen2007]. Excavations revealed abundant trade goods, including imported ceramics and stonewares, metal artifacts, glass beads, and stone beads, indicative of frequent trade activities. The introduction of foreign prestige goods to local Indigenous communities might have occurred before European presence became established, but was maintained and amplified after European arrival due to local competition for power [@Borao2009; @LiandWu2006; @LWandBM2020ornament; @Ueda2016]. A general model to summarize this situation in northeastern Taiwan might assume that the influence of a colonial power, combined with high local values attached to imported goods, might lead to increased social inequality due to competition among individuals [@Brumfiel1994; @Clark1994]. The observed uneven distribution in Northeastern Taiwan may be explained as a result of unequal access to trade goods when Indigenous societies became more involved in the complex trade network stimulated by Europeans.

@@

## Exponential random graph models in a Bayesian framework  {-}

<!-- ERGM in archaeology, and limitations --> 
Exponential random graph models (ERGMs) first appeared in archaeology with @Brughmans2014, who studied Iron age settlement patterns in Southern Spain by modeling inter-settlement visibility networks and visual control at 159 sites. By fitting models with the observed network of archaeological data, they proposed that ERGMs is a promising method for exploration and hypothesis testing for social processes. Similarly, @Amati2019 modeled three networks consisting of 15 sites (AD 100 to 400) in the Caribbean to explore interaction mechanisms, including proximity, inter-cultural items, and pottery types. By comparing hypothesized networks with the observed sites, they found that the presence of hub sites can be efficiently explained by multiple interdependent mechanisms instead of only one variable exclusively. However, those studies also point out some limitations of ERGMs, such as sensitivity to missing data and less able to handle uncertainty. Also, it is difficult in ERGMs to estimate model parameters and interpret the result due to intractable likelihood normalizing constants and model degeneracy [@Caimo2014; @Jin2013]. A normalizing constant is a function of the model parameter for making probability distributions integrate to one, which becomes harder to compute with a larger set of networks [@Caimo2017bayesian]. This is also termed "doubly intractable" since both the likelihood normalizing constant and the marginal likelihood (the evidence of the posterior) are hard to derive [@Caimo2013; @Lyne2015]. Model degeneracy is another issue where probability models tend to overestimate a small number of extreme graphs by assigning too much weight, e.g., in extreme cases such as empty (all nodes unconnected) or complete graphs (all nodes connected) [@Schweinberger2011; @Caimo2014]. One solution to these limitations is available by implementing ERGMs by means of a Bayesian framework.

@@

<!-- Bayesian inference for Exponential Random Graph Models --> 
Bayesian approaches to ERGMs are effective tools for network modeling by incorporating prior information about the network configurations to better understand dependencies of network variables and improve computational issues in ERGMs [@Caimo2017; @Lehmann2020]. Prior information is derived from previous data or assumptions of our data. The advantage that Bayesian modeling has over traditional ERGMs is the application of Markov chain Monte Carlo (MCMC) simulation using the approximate exchange algorithm [@Caimo2011]. With the exchange algorithm, Bayesian ERGMs avoid doubly-intractable computations by directly sampling from the not normalized part of the posterior, which alleviates the computational problems and gives better convergence results. This enables us to deal with complicated dependence patterns with ease, providing better estimations for complex social network models with heterogeneous data [@Caimo2017; @Snijders2006]. By fitting an ERMG with the approximate exchange algorithm, a Bayesian approach generates posterior probabilities that incorporate our sample data and prior information that estimates the effect of each ERGM parameter in our models [@Caimo2015; @Nemmers2019]. This allows uncertainty quantification by examining the posterior mean and 95% credible intervals [@Caimo2017bayesian]. The typical criteria for interpreting a significant effect of a parameter is that its credible interval does not contain zero [@Harris2013, pp. 51]. In addition, Bayesian approaches are useful to deal with missing data, which is often a problem leading to misinterpretation of networks, especially for archaeological studies. @Koskinen2010 shows that the effect of missing data can be reduced with Bayesian modeling that can predict, on the average, 80% of the ties when a third of data is missing.

@@

<!-- archaeological setting -->
We analyzed burial data collected from the excavation reports, and the original fieldwork notes for the upper component of Kiwulan (1350-1850 AD) (Figure \@ref(fig:figure-KWL-map)) [@Chen2007]. A total of 90 burials were unearthed from adjacent squares that provide continuous stratigraphic sections suitable for temporal comparison (Figure \@ref(fig:figure-burial-map)). Burials are oriented in an east-west direction on the north side of the residential area, indicated by post-holes and _in-situ_ wooden posts, suggesting a well-organized spatial arrangement. Previous studies report an uneven distribution of prestige goods across burials without agreement about whether this uneven distribution hints at vertical social differences. For example, @Cheng2008 interpreted the unequal distribution of glass beads, especially the gold-foil beads between burials, as evidence for hierarchy, indicating a stratified society. However, @Hsieh2012 suggested a relatively egalitarian structure based on the comparative analysis of the frequencies of all burial goods. She found that the burials with rare prestige goods were usually associated with elders, which might indicate achieved, rather than inherited, status. One important limitation of these previous studies is that they did not use analytical units suitable for comparing the phases before and after European arrival. Here, we adopt a new chronological framework for the burials to test if network configurations differ from the pre-European period (before European colonization) to the post-European period (during and after European colonization until Han Chinese immigrants arrival). The discord over the discussion of Kiwulan social organization could be associated with chronological differences that have not yet been well studied.

@@

<!-- burial data and the definition of tie formation  --> 
We used burial goods in common to construct ties for representing a flow of goods between burials. Since trade beads are commonly found across burials with substantial differences in quantities, we described each burial as having one of four levels, high, upper-middle, lower-middle, and low, according to their distributions across all burials (details in Supplementary Online Materials). For less frequent prestige goods, including imported ceramics, gold foils, and fish-shaped ornaments, we linked two burials when they possess each type of goods (i.e. presence or absence). Node attributes here include osteological data, such as age and sex, and cultural data, such as ritual pottery, and burial value index. Ritual pottery was identified as locally made ceramics placed above graves that suggests funeral feasting based on historical records. The burial value index is an important attribute for economical inequality that we assigned the burials into three ranks, high (>`r high_value`, top 10 percent), above average (`r ave_value`-`r high_value`), and below average (<`r ave_value`), as an index of wealth. The number is the sum of values of each type of prestige good, which is calculated by the total number of prestige items from burial context over the number of prestige items in a burial. Since burials tend to have multiple prestige goods in common, the network ties are weighted instead of binary data (the value 1 represents a tie and the value 0 otherwise) [@Snijders2011]. For example, if two burials have both low quantity of glass beads and porcelain in common, the tie is given a value of 2. Our networks are non-directed, which means a mutual relationship where the tie between any two actors is bidirectional. The networks constructed based on these principles show that the network after the European presence has more node connectivity in general with some nodes having a larger number of connections (Figure \@ref(fig:figure-network-diagrams)).

@@

<!-- Network construction and specification -->
We quantify the relations among burials and test our hypothesis of social change using the R programming language [@Rlanguage2019] with the bergm package [@Caimo2014], where model specification and simulation procedures is based on statnet package [@Hunter2008; @Morris2008; @Handcock2008]. Table \@ref(tab:table-ergm-terms) lists the network parameters we used for the model specification. Every parameter in an ERGM has an associated algorithm for computing the probability of observing relations in terms of grave goods between two burials. Based on our hypothesis, we model a network with increased social inequality to be represented by endogenous network effects: low transitivity and high popularity. We include burial-specific attributes as covariate effects for homophily, such as age, sex, ritual activity, and the degree of wealth, to explore the tendency of tie forming. For example, age-homophily means people of the same age tend to have the same burial goods. We also include the physical distance between burials as an indicator of a kinship-based relation since the deceased from the same family tend to be buried nearby [@LiandWu2006]. Our model could reveal the emergence of social inequality via the presence of a few individuals as network centers, having more relations with others. This would be indicated by high popularity or degree values in the network statistics, with covariates to understand the preferential tendency of formation of relationships. According to our hypothesis, the burial evidence from after the European arrival will show higher popularity compared with the burial evidence from before European contact.

@@

<!--posterior stats for for post-E and their comparison-->
For the network model after the arrival of Europeans, the nodal covariates of wealth rank show significant positive effects, while ritual and sex variables have negative effects. This indicates the burials in the same wealth level tend to have relations. On the contrary, the same sex burials and burial with ritual pottery in common tend not to have relations. There are no significant effects for variables of age and physical distance. Similar to the pre-European network, the endogenous network variable, transitivity (gwesp) demonstrates a significant positive effect but much lower than the effect of pre-European network. In contrast, the popularity (gwdegree) shows a significantly higher positive effect than the effect of the pre-European network. This means there is a tendency toward centralization that reflects a limited number of burials having many more ties than others. This implies that the presence of better access to trade goods and the behavior of wealth accumulation or display in burial events. In general, the post-European network model has a smaller transitivity effect and a positive popularity effect than the pre-European network model. This may suggest a reduced tendency toward clustering but high tendency toward centralization after the European presence. In addition, both posterior estimates present symmetric distributions, with the posterior means close to posterior medians (Table \@ref(tab:table-posterior-stats)).

@@

<!-- comparison between two networks using bootstrap --> 
One key difference between the pre-European and post-European networks is their size, with `r pre_E` burials compared to `r post_E` burials. To understand the robustness of comparison between two networks, we used vertex bootstrap to cross-validate the results of Bayesian ERGMs. Vertex bootstrap is a non-parametric method that conducts resampling for all vertices (i.e. node) that is useful for quantifying standard errors and estimating sampling variability in the network statistics of interest [@Chen2019; @Snijders1999; @Rroberts2021]. This enables the evaluation of uncertainty for networks and tests the difference between multiple networks by examining their confidence intervals for the network population. Networks with nonoverlapped intervals means there is a significant difference. We computed and compared endogenous network statistics, including density, popularity, and transitivity for our two networks. Figure \@ref(fig:figure-bootstrap-CI) shows a significant difference in observed network popularity between the two networks, which is consistent with our finding of negative popularity in the pre-European period and positive popularity in the post-European period using the Bayesian approach. For density and transitivity, the 95% confidence intervals from the two networks overlapped with each other that indicates no significant difference. This is also consistent with our results of Bayesian modeling, where both networks present similar positive or negative effects with some degree of difference.

@@

<!--archaeological implication based on the results-->
The covariants for networks provide further information to give insights into how specific elements of the archaeological record relate to the formation of networks (Table \@ref(tab:table-posterior-stats)). In other words, we can know what kinds of socio-cultural factors associated with the acquisition and distribution of foreign prestige goods reflected by burial networks. Our covariants were age, sex, ritual pottery, and wealth ranks calculated based on the value of grave goods. The results indicate that the major factor contributing to ties between burials is ritual for the pre-European network, and wealth for the post-European network. Moreover, in the post-European network, burials with ritual vessels or same sex tend not to be tied that suggest they are not associated with the network pattern we observed. Those results illustrate that different social and economic mechanisms determined relations between burials, or the interconnections of prestige goods, for the two periods. Before European arrival, ritual behaviors seem to have been important factors that structure the interconnections of prestige goods in the burial context. This might imply that ritual could be a status indicator when the foreign goods were first introduced to the local society. After European presence was established, status indicators shifted from ritual to wealth differences with the least sex status discrimination. Moreover, geographical proximity, implying some kinship association, is not a factor for sharing similar prestige items between burials for both phases. This may suggest that the use of prestige items is more individual-oriented instead of kinship-based. Demographic information, such as age and sex, do not have any effects on network formation for the pre-European network and only sex shows a negative effect for the post-European network. This may also support a shift to more wealth and individual-based status competition without sex differences in the later phase. But we also note that no effects may be due to the unavailability of age and sex values for many burials in our sample.

<!-- other evidence -->
We used burial data as a proxy to explore social relations reflected by interconnections of prestige goods based on the assumption that burials can represent social structures [@Binford1971; @Saxe1970]. We argue that the social relation at Kiwulan changed because of the contact with European colonial power and associated foreign trade. This argument is supported by the changes in burial network structures from before and after European arrival and their associated underlying mechanisms. It should be noted that manipulation of burial rituals by the living can cause a disconnect between a person's status in life and their status represented by burial contexts [@Hodder1980; @Pearson1982]. However, this issue can be reduced by comparing the results with evidence from other archaeological contexts, such as residential areas [@Chapman2003]. Previous studies of trade ornaments from the residential area of Kiwulan suggests an uneven spatial distribution when the Europeans were active in northern Taiwan that hints at an increased social differentiation at Kiwulan in a pericolonial situation [@LWandBM2020ornament]. This is consistent with the result of a more centralized burial network after the presence of Europeans, which supports a connection between a burial and the social role of the living person. Since the burial goods used for making ties were treated as prestige goods or status items in the local Indigenous culture [@Borao2009; @LiandWu2006], the interrelations represented by the flow of goods observed in the burial data likely reflect social contexts of the living people at Kiwulan in a pericolonial context.

@@

We would like to thank the Yilan County Cultural Affairs Bureau in Taiwan for permitting access to the data used in this paper. We thank the excavation team of Kiwulan led by Yu-pei Chen for their post-excavation analysis and detailed documentation about burials that enable this study. We also thank Alberto Caimo and Steven M. Goodreau for the discussion regarding the methods at the earliest stage of analysis. This work was supported in part by the travel grant from the Department of Anthropology at the University of Washington and the Doctoral Dissertation Fellowship from the Chiang Ching-kuo Foundation for International Scholarly Exchange (Project #DF009-A-18). We thank Ben Fitzhugh and Peter Lape for their insightful comments on an early draft. We also acknowledge comments from anonymous reviewers that have greatly improved this manuscript.
